Modelling multi-scale state-switching functional data with hidden Markov models

### 2021-05-22 Report/Sections/models.tex
update reviewer 1, comments 1-4a

%Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
%Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be synthesized to form a wide variety of new models.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HMMs as a base structure}
\label{subsec:HMM}

@@

Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$. The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
%

Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to 1:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability. This simplifies likelihood maximization by removing constraints in the optimization problem. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$. Figure \ref{fig:models}a represents the dependence structure of an HMM.

@@

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$. A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
%
\begin{align}
\label{eqn:carhmm}
\begin{split}
\mathbb{E}(Y_{t}|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \\
\mathbb{V}(Y_t| Y_{t-1} = y_{t-1},X_t=i) &= (\sigma^{(i)})^2.
\end{split}
\end{align}
%
A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.

As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. If $y$ is the sequence of observed emissions, then
\begin{equation*}
    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
    \label{eqn:CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$. Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.

@@

A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs. The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $\{X_1, \ldots, X_T\}$ is an unobserved Markov chain with $N$ possible states and $\{Y_1, \ldots, Y_T\}$ is the set of corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.   
%
In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each curve $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$. If $X_t = i$, then the components of $X_t^*$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain. The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$. We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\big\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\big\}$. In summary:
%
% this is new
%
\begin{gather*}
    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
    %
    (Y_t | Y_{t-1} = y_{t-1}, X_t = i) \text{ has density } f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
    %
    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
    %
    (Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i) \text{ has density } f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
\end{gather*}
Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
%
% end new
%

Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM. For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$. We also share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.

Due to the nested structure of the HHMM, the likelihood is easily calculated using the forward algorithm.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N,
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models. If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ both have $(i,i)^{th}$ entries equal to 
$f^{(i)}(y_t)\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$. 
If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$.
The fine-scale likelihood $\calL_{\text{fine}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
%
%For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.

@@

In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM-based models described thus far. 
%For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window over the elements of $Y_t^*$. Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window. Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,h+1}$, and so on. Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting. In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$. Namely:
%
\begin{align}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
    \label{eq:DFTdef}
\end{align}
%
for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i = \sqrt{-1}$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. We omit the final window if $t^*+h-1$ exceeds $T^*_t$, denote the total number of windows as $\tilde T^*_t$, and index each window with $\tilde{t}^* = 1,\ldots,\tilde T^*_t$. Next, we calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|_2^2,
\end{equation}
%
where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window. In words, $\Zone_{t,\tilde{t}^*}$ is the average value of $Y^*_t$ within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window. %(by the Plancherel theorem). 
More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the fine-scale data within curve $t$ and window $\tilde{t}^*$.

When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. As a result, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$. The likelihood of this model is therefore identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$. To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT. Figure \ref{fig:models}d displays the dependence structure of an HMM-DFT.

>>>>

\subsection{HMMs as a base structure}
\label{subsec:HMM}

@@

Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$. The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:

\[\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,\]
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
%

Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to 1:

\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability. This simplifies likelihood maximization by removing constraints in the optimization problem. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$. Figure \ref{fig:models}a represents the dependence structure of an HMM.

@@

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$. A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

\begin{align}
\label{eqn:carhmm}
    \begin{split}
    \mathbb{E}(Y_{t}|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \\
    \mathbb{V}(Y_t| Y_{t-1} = y_{t-1},X_t=i) &= (\sigma^{(i)})^2.
    \end{split}
\end{align}
%
A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.

As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. If $y$ is the sequence of observed emissions, then

\begin{equation*}
    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
    \label{eqn:CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$. Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.

@@

A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs. The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $\{X_1, \ldots, X_T\}$ is an unobserved Markov chain with $N$ possible states and $\{Y_1, \ldots, Y_T\}$ is the set of corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.   
%
In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X^*_t \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$. If $X_t = i$, then $X^*_{t,t^*} \in \{1,\ldots,N^{*(i)}\}$ represents one of $N^{*(i)}$ possible within-curve behaviours associated with curve type $i$. In turn, $X^*_t$ emits a sequence of fine-scale observations $Y^*_t \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each curve $t$, the fine-scale process $\{X^*_t, Y^*_t\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$. If $X_t = i$, then the components of $X^*_t$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain. The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$. We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\big\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\big\}$. In summary:

\begin{gather*}
    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
    %
    (Y_t | Y_{t-1} = y_{t-1}, X_t = i) \text{ has density } f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
    %
    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
    %
    (Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i) \text{ has density } f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
\end{gather*}
Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X^*_T, Y^*_T\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.

Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM. For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$. We also share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, \textit{not} their fine-scale emission distributions.

While at first it seems difficult to jointly infer the paramtersDue to the nested structure of the HHMM, its likelihood is easily calculated using the forward algorithm.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then

\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N,
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models. If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ both have $(i,i)^{th}$ entries equal to 
$f^{(i)}(y_t)\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$. 
If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$.
The fine-scale likelihood $\calL_{\text{fine}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.

@@

In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM-based models described thus far. To handle these additional fine-scale structures, we recommend replacing $Y^*_t = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with stride length $h$ over the elements of $Y^*_t$. Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,1+h}$, and so on.
Replacing $Y^*_{t,t^*}$ with relevant summary statistics results in a loss of information since we are substituting infinite-dimensional curves with a low-dimensional representation. \citet{Aue:2015} and \citet{Gao:2019} use functional principle component analysis (FPCA) for dimension reduction to minimize error when predicting functional data. However, these papers do not account for state-switching behaviour between curves and do not guarantee that the resulting FPCA scores will be interpretable for practitioners. The goal of this paper is to classify and describe curve types and within-curve behaviour, not predict future functional data. We are thus primarily interested in the fine-scale summary statistics themselves rather than the curves from which they are calculated, and these summary statistics are known to have consistent maximum likelihood estimates \cite{Douc:2004}. As a result, we choose a dimension reduction technique which can effectively separate behavioural states and has an intuitive interpretation for practitioners.
In general, subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window. Larger stride lengths result in a larger loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting. In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows. For our particular case study, we use the discrete Fourier transform (DFT) of a moving forward window with width $h$ and stride $h$ across $Y^*_t$:

\begin{align}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
    \label{eq:DFTdef}
\end{align}
%
for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i = \sqrt{-1}$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. We omit the final window if $t^*+h-1$ exceeds $T^*_t$, denote the total number of windows as $\tilde T^*_t$, and index each window with $\tilde{t}^* = 1,\ldots,\tilde T^*_t$. Next, we calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:

\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|_2^2,
\end{equation}
%
where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window. In words, $\Zone_{t,\tilde{t}^*}$ is the average value of $Y^*_t$ within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window. More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the fine-scale data within curve $t$ and window $\tilde{t}^*$.

When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. As a result, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$. The likelihood of this model is therefore identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$. To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT. Figure \ref{fig:models}d displays the dependence structure of a fine-scale HMM-DFT.
